---
title: "M3: Class Exercise on Trend and Season"
author: "Luana Lima"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output. 

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)

```

## Importing data

Let's continue working with our inflow data for reservoirs in Brazil. 

```{r}

#Importing time series data from text file#
raw_inflow_data <- read.table(file="./Data/inflowtimeseries.txt",header=FALSE,skip=0)

#Trim the table to include only columns you need
nhydro <- ncol(raw_inflow_data)-2
nobs <- nrow(raw_inflow_data) 

#If your file does not have header like this one you can add column names after 
#creating the data frame
colnames(raw_inflow_data)=c("Month","Year", "HP1", "HP2","HP3","HP4", "HP5",
                            "HP6","HP7", "HP8","HP9","HP10", "HP11","HP12", 
                            "HP13", "HP14","HP15")
#Checking data
head(raw_inflow_data)
str(raw_inflow_data)
```
## Creating the date object

Here we use the function my() from package lubridate.

```{r}
#using package lubridate
my_date <- paste(raw_inflow_data[,1],raw_inflow_data[,2],sep="-")
my_date <- my(my_date)  #function my from package lubridate
head(my_date)
#add that to inflow_data and store in a new data frame 
inflow_data <- cbind(my_date,raw_inflow_data[,3:(3+nhydro-1)])
head(inflow_data)
```


## Initial Plots

Initial time series plot.

```{r}
#using package ggplot2
for(i in 1:nhydro){
  print(ggplot(inflow_data, aes(x=my_date, y=inflow_data[,(1+i)])) + 
          #1+i because of the column data, the first column is the date and we want to keep track of indexes
          #need the print function to print the plot within a for loop! 
            geom_line(color="blue") +
            ylab(paste0("Inflow ",colnames(inflow_data)[(1+i)],sep="")) 
        )
}

```

## Zeros in the end on data 

The initial plots showed that we have zeros in the end of the data set. It could be missing observation or observation that haven't been observed yet. Use the tail() to find out how many zeros you have and how many lines you will need to remove.

```{r}
#check the final obs on data 
tail(inflow_data)


```

Note our last observation is from August 2011 but the data file was filled with zeros. Let's remove the last four rows of our data set.

```{r}
#Remove last year by replacing current data frame
inflow_data <- inflow_data[1:(nobs-12),]

#update object with number of observations (in order to have full cycles of data for each year)
nobs <- nobs-12

#Tail again to check if the rows were correctly removed
tail(inflow_data)
```
Fixed!

## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*. 

```{r}
ts_inflow_data <- ts(inflow_data[,2:(2+nhydro-1)],start= c(1931,1),frequency=12) 
#if you dont specify start, you don't get the year your data begins
#note that we are only transforming columns with inflow data, not the date columns  #start=my_date[1],end=my_date[nobs],
head(ts_inflow_data,15)
```


## Plotting ACF and PACF

Let's use functions Acf() and Pacf() from package "forecast".

```{r acf}
#Acf and Pacf for HP1
for(i in 1:nhydro){
  par(mfrow=c(1,2))  #place plot side by side with par, 1 row and 2 columns 
  Acf(ts_inflow_data[,i],lag.max=40,main=paste("Inflows HP",i,sep="")) 
  # because I am not storing Acf() into any object, I don't need to specify plot=TRUE 
  Pacf(ts_inflow_data[,i],lag.max=40,main=paste("Inflows HP",i,sep=""))
}
#example in #4, there is likely a strong trend component because the ACF is steadily decreasing over time.. (unsure if there is a seasonal component)
```


## Trend Component - Linear Model

Let's identify and remove trend component. You start by fitting a linear model to $Y_t=\beta_0+\beta_1*t+\epsilon_t$. 

```{r}
#Create vector t
t<- 1:nobs #or could use mydate (but makes interpretation a little more confusing)

#for (i in 1:nhydro){} to do this as a for loop with iHP defined as i....
iHP <- 4 #hydro plant 4
#Fit a linear trend to TS of iHP
linear_trend <- lm(inflow_data[,(iHP+1)]~t) #inflow is increasing for every unit of time by 2.18 (strong correlation)
#but only 26% of the variability is caused by time, so there is more influencing your data than just time
#but we do see that the trend is signifcant 
summary(linear_trend)

```


```{r}
#remove the trend from series
beta0 <- linear_trend$coefficients[1]#intercept
beta1 <- linear_trend$coefficients[2]#slope

#original ts, minus the trend defined above from lm(), the epsilon is left in the data and is part of the random component of the time series (y_detrend is the epsilon (whatever was not estimated by the linear component))
y_detrend <- inflow_data[,(iHP+1)]- (beta0 +beta1*t)

df_detrend <- data.frame("date"= inflow_data$my_date, 
                         "observed" = inflow_data[,iHP+1], 
                         "detrend"=y_detrend)

ggplot(df_detrend, aes(x=date)) +
  geom_line(aes(y=observed), color="tomato4") +
  geom_line(aes(y=detrend), color= "forestgreen")
 # geom_abline(beta0,beta1)
```

Note that blue line is our original series, red line is our trend, green line is our original series minus the trend or in other words the detrended series. And in orange is the trend line for the detrended series which has slope 0 meaning we were able to effectively eliminate the trend with a linear model.

#Check what happened to ACF after removing the linear trend

```{r}
library(cowplot)
#install.packages("ggfortify")
library(ggfortify)

#making a timeseries of the detrended data
ts_y_detrend <- ts(y_detrend, start= c(1931,1),frequency=12)


plot_grid(
  autoplot(ts_inflow_data[,iHP],ylab="Observed Inflow"),
  autoplot(Acf(ts_inflow_data[,iHP],lag.max=40,plot=FALSE),main=NULL),
  autoplot(ts_y_detrend,ylab="Observed Inflow - Linear Model"),
  autoplot(Acf(ts_y_detrend,lag.max=40,plot=FALSE),main=NULL),
  nrow=2,ncol=2
)
#see that after removing the trend, the autocorrelation goes down (want less time dependence, so that we need less elements in order to model future data)

#want a model that represents out datasets and is a combination of other variables, how many lagged observations do we need to add? want less parameters (less error for each parameter), and want time dependence represented in initial lags instead of the larger lags (could write TS with only 12 terms, and the trend term)

#look at pacf to find the value of the coefficients that we are going to add to our model


```


## Seasonal Component

Now let's shift attention to the seasonal component.

```{r}
#Use seasonal means model
iHP <- 1 #this plant seems to have strong seasonal trend

#create dummies
dummies= seasonaldummy(ts_inflow_data[,iHP]) #seasonal dummy must use a ts object

#linear model where our original data is the y dependent upon dummies 
seas_linear_model <- lm(inflow_data[,(iHP+1)]~dummies) #iHP+1 because of the date column
summary(seas_linear_model)
#april is insignificant because it isn't that different from the intercept, December
#in results, the intercept term is december, if you want dummiesdec, then you must run the linear model without the intercept term
#r2 tells us how much variability is caused by the seasonal relationship 

#storing coef
beta0 <- seas_linear_model$coefficients[1]
beta1 <- seas_linear_model$coefficients[2:12]

#seasonal component to subtract from the original dataset
seas_comp <- array(0,nobs)
for(i in 1:nobs){
  seas_comp[i] <- beta0 + beta1 %*% dummies[i,] #element by element for row i (beta1 times dummy for feb... march... etc)
}
```

#visualizing above
```{r}
ggplot(inflow_data, aes(x=my_date))+
  geom_line(aes(y=inflow_data[,(iHP+1)]), col="black")+
  geom_line(aes(y=seas_comp),col="orange")

y_deseason <- inflow_data[,(iHP+1)] - seas_comp

ggplot(inflow_data, aes(x=my_date))+
  geom_line(aes(y=inflow_data[,(iHP+1)]), col="black")+
  geom_line(aes(y=y_deseason),col="green")
```

## Check what happened to ACF after removing the linear trend 

```{r}
ts_y_deseason <- ts(y_deseason, start= c(1931,1),frequency=12)


plot_grid(
  autoplot(ts_inflow_data[,iHP],ylab="Observed Inflow"),
  autoplot(Acf(ts_inflow_data[,iHP],lag.max=40,plot=FALSE),main=NULL),
  autoplot(ts_y_deseason,ylab="Observed Inflow - Linear Model"),
  autoplot(Acf(y_deseason,lag.max=40,plot=FALSE),main=NULL),
  nrow=2,ncol=2
)
#we still see some seasonality in the acf, thats because we are using a constant seasonal model, and there could be some shifts in the sesaonal in the data actually... 

#seasonal component is not well pronounced, or it is still there... 
```
## decomposing a timeseries
```{r}
iHP <- 1
plot(decompose(ts_inflow_data[,iHP]))#only accepts time series objects
```


##Exercise

Fit trend and seasonal for the other variables HP2, HP3, ...


## Stationarity Tests in R

Some test only work for non-seasonal data. So let's create another series with yearly averages for inflow.

```{r}
#Group data in yearly steps instances


```

### Mann Kendall 

Check for deterministic trend.

```{r}

#Since I have seasonal data I cannot use the simple MannKendall()
#another example of functions that need a ts object


#Use yearly date to run Mann Kendall

```

### Spearman Correlation test 

Also check for deterministic trend, for non-seasonal data only.

```{r}
#Deterministic trend with Spearman Correlation Test
print("Results from Spearman Correlation")


#with cor.test you can get test statistics

```

### Augmented Dickey Fuller (ADF)

Used to detect stochastic trend.

```{r}
#Null hypothesis is that data has a unit root
print("Results for ADF test/n")

#Now let's try the yearly data
print("Results for ADF test on yearly data/n")

```

##Exercise

Run the stationarity tests for the other variables HP2, HP3, ...

