---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 2 - Due date 02/25/24"
author: "Emma Kaufman"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r, load packages}
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
library(lubridate)
library(readxl)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2023 Monthly Energy Review. The spreadsheet is ready to be used. You will also find a $.csv$ version of the data "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv". You may use the function $read.table()$ to import the $.csv$ data in R. Or refer to the file "M2_ImportingData_CSV_XLSX.Rmd" in our Lessons folder for functions that are better suited for importing the $.xlsx$.

```{r, importing data}
#Importing data set without change the original file using read.xlsx
energy_data1 <- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE) 

#Now let's extract the column names from row 11
read_col_names <- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE) 

colnames(energy_data1) <- read_col_names
head(energy_data1)

```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r, selecting desired data}
#selecting desired columns
energy_interest <- select(energy_data1, 'Total Biomass Energy Production', 'Total Renewable Energy Production', 'Hydroelectric Power Consumption')

#previewing new dataframe
head(energy_interest)
```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r, TS object}
#creating a timeseries object
energy_ts <- ts(energy_interest, start = c(1973,1), frequency = 12)

```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
#extracting each time series 
biomass <- energy_ts[,1]
renewable <- energy_ts[,2]
hydro <- energy_ts[,3]

#computing mean
mean_bio <- mean(biomass)
mean_renewable <- mean(renewable)
mean_hydro <- mean(hydro)

#computing std dev
sd_bio <- sd(biomass)
sd_renewable <- sd(renewable)
sd_hydro <- sd(hydro)


```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}
#Biomass_Decomposed <- stl(biomass, s.window = "periodic") 
#plot(Biomass_Decomposed)
plot(biomass) + abline(h= mean_bio, col = "green")

plot(renewable) + abline(h= mean_renewable, col = "orange")

plot(hydro) + abline(h= mean_hydro, col = "blue")

```


## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}

```


## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}

```


## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}

```


