---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 3 - Due date 02/01/24"
author: "Emma Kaufman"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
library(lubridate)
library(readxl)
library(ggplot2)
library(cowplot)
library(gridExtra)
```

```{r}
#Importing data set without changing the original file using read.xlsx
energy_data1 <- read_excel(path=
                        "./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
                          skip = 12, 
                          sheet="Monthly Data",
                          col_names=FALSE) 

#Now let's extract the column names from row 11
read_col_names <- read_excel(path=
                          "./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
                             skip = 10,
                             n_max = 1, 
                             sheet="Monthly Data",
                             col_names=FALSE) 

colnames(energy_data1) <- read_col_names
```

```{r}
#creating a dataframe of the columns we are interested in
energy_interest <- select(energy_data1, 
                          'Total Biomass Energy Production', 
                          'Total Renewable Energy Production', 
                          'Hydroelectric Power Consumption')
#creating a timeseries object
energy_ts <- ts(energy_interest, start = c(1973,1), frequency = 12)
biomass_ts <- energy_ts[,1]
renewable_ts <- energy_ts[,2]
hydro_ts <- energy_ts[,3]
```


##Trend Component

### Q1
For each time series, i.e., Renewable Energy Production and Hydroelectric Consumption create three plots: one with time series, one with the ACF and with the PACF. You may use the some code form A2, but I want all the three plots side by side as in a grid.
(Hint: use function `plot_grid()` from the `cowplot` package)

###Renewable Energy Production
```{r, q1a}
#time series plot
Renew_ts_plot<- autoplot(renewable_ts) +
  labs(x = "Time", 
       y = "Total Renewable \nEnergy Production \n(trillion BTU)", 
       title = "Renewable Energy Production over Time")

#acf plot
Renewable_acf= autoplot(Acf(renewable_ts,lag.max=40))

#pacf plot
Renewable_pacf= autoplot(Pacf(renewable_ts,lag.max=40))

#plotting all three as grid
plot_grid(Renew_ts_plot,Renewable_acf, Renewable_pacf, nrow=3,align= 'h', rel_heights = c(2, 2, 2))


```

###Biomass Energy Production
```{r, q1b}
#time series plot
Biomass_ts_plot<-autoplot(biomass_ts) +
  labs(x = "Time", 
       y = "Total Biomass \nEnergy Production \n(trillion BTU)", 
       title = "Biomass Energy Production over Time") 

#acf plot
Biomass_acf= autoplot(Acf(biomass_ts,lag.max=40))

#pacf plot
Biomass_pacf= autoplot(Pacf(biomass_ts,lag.max=40))

#plotting all three as grid
plot_grid(Biomass_ts_plot,Biomass_acf, Biomass_pacf, nrow=3,align= 'h', rel_heights = c(2, 2, 2))

```

###Hydroelectric Consumption
```{r, q1c}
#time series plot
Hydro_ts_plot<-autoplot(hydro_ts) +
  labs(x = "Time", 
       y = "Total Hydroelectric \nPower Consumption \n(trillion BTU)", 
       title = "Hydroelectric Power Consumption over Time")

#acf plot
Hydro_acf= autoplot(Acf(hydro_ts,lag.max=40))

#pacf plot
Hydro_pacf= autoplot(Pacf(hydro_ts,lag.max=40))

#plotting all three as grid
plot_grid(Hydro_ts_plot,Hydro_acf, Hydro_pacf, nrow=3,align= 'h', rel_heights = c(2, 2,2))


```

### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

>Hydroelectric Power Consumption does not appear to have a trend. It oscillates above and below the mean consistently over time, and the mean appears to be consistent. Biomass and energy production both have upwards trends. They both follow a somewhat polynomial upwards trend, that could be broken up into piecewise linear trends. 

### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r}
nobs <- nrow(energy_interest)
#Create vector t
t<- 1:nobs

beta0 <- numeric(3)
beta1 <- numeric(3)
energy_interest<- as.data.frame(energy_interest)

for (i in 1:3) {
  linear_trend <- lm(energy_interest[, i] ~ t) #define linear trend based on interested energy type
  print(summary(linear_trend))
  beta0[i] <- coef(linear_trend)[1]  # intercept
  beta1[i] <- coef(linear_trend)[2]  # slope
}
beta0
beta1
```
The slopes stored in beta1 represent how much energy production/ consumption is increasing for each time step. For Biomass Energy Production the p-value is <0.05 so the slope of 0.477 is significant, and the r2 of 0.82 means that this linear trend accounts for 82% of the variability in the data. 
For Total Renewable Energy Production the p-value is also <0.05 so the slope of 0.7 is significant, and the r2 of 0.807 means that this linear trend accounts for 80.7% of the variability in the data. 
For the Hydroelectric Power Consumption the p-value is <0.05 so the slope of -0.009 is significant (though the magnitude is very small). But the r2 is 0.01, so this linear trend accounts for only 1% of the variability in the data. 
*Add in explanation about the intercepts??* 

### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?
```{r}
#Total Biomass
i=1
  y_detrend_bio <- energy_interest[, i] - (beta0[i]+beta1[i]*t)
  df_detrend_bio <- data.frame("date"=energy_data1$Month,
                           "observed"=energy_interest[, i],
                           "detrend"= y_detrend_bio)
  detrended_series_bio <- ggplot(df_detrend_bio, aes(x=date))+
  geom_line(aes(y=observed), color="tomato4")+
  geom_line(aes(y=detrend), color= "forestgreen") +
  ggtitle(paste("Detrended Series (in green) for", colnames(energy_interest)[i])) +
    labs(y= "Energy (trillion BTU)")
  plot(detrended_series_bio)
  
#Total Renewable
  i=2
  y_detrend_renew <- energy_interest[, i] - (beta0[i]+beta1[i]*t)
  df_detrend_renew <- data.frame("date"=energy_data1$Month,
                           "observed"=energy_interest[, i],
                           "detrend"= y_detrend_renew)
  detrended_series_renew <- ggplot(df_detrend_renew, aes(x=date))+
  geom_line(aes(y=observed), color="tomato4")+
  geom_line(aes(y=detrend), color= "forestgreen") +
  ggtitle(paste("Detrended Series (in green) for", colnames(energy_interest)[i])) +
    labs(y= "Energy (trillion BTU)")
  plot(detrended_series_renew)

#Hydroelectric
  i=3
  y_detrend_hydro <- energy_interest[, i] - (beta0[i]+beta1[i]*t)
  df_detrend_hydro <- data.frame("date"=energy_data1$Month,
                           "observed"=energy_interest[, i],
                           "detrend"= y_detrend_hydro)
  detrended_series_hydro <- ggplot(df_detrend_hydro, aes(x=date))+
  geom_line(aes(y=observed), color="tomato4")+
  geom_line(aes(y=detrend), color= "forestgreen") +
  ggtitle(paste("Detrended Series (in green) for", colnames(energy_interest)[i])) +
    labs(y= "Energy (trillion BTU)")
  plot(detrended_series_hydro)

```
The detrended series all are lower magnitude


### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. but not mandatory. Did the plots change? How?
```{r}
#creating timeseries objects for all of the detrended series
bio_detrend_ts <- ts(y_detrend_bio, start = c(1973,1), frequency = 12)
renew_detrend_ts <-ts(y_detrend_renew, start = c(1973,1), frequency = 12)
hydro_detrend_ts <- ts(y_detrend_hydro, start = c(1973,1), frequency = 12)
```

```{r, bio}
#acf detrended plot for biomass
bio_acf_detrend= autoplot(Acf(bio_detrend_ts,lag.max=40))

#pacf detrendedplot for biomass
bio_pacf_detrend= autoplot(Pacf(bio_detrend_ts,lag.max=40))

#plotting comparison with plots from #1 as grid
plot_grid(Biomass_acf, bio_acf_detrend,Biomass_pacf, bio_pacf_detrend, nrow=4,align= 'h', rel_heights = c(2, 2,2,2))
```

```{r, renewables}
#acf detrended plot for renewables
Renew_acf_detrend= autoplot(Acf(renew_detrend_ts,lag.max=40))

#pacf detrendedplot for renewables
Renew_pacf_detrend= autoplot(Pacf(renew_detrend_ts,lag.max=40))

#plotting comparison with plots from #1 as grid
plot_grid(Renewable_acf, Renew_acf_detrend,Renewable_pacf, Renew_pacf_detrend, nrow=4,align= 'h', rel_heights = c(2, 2,2,2))
```

```{r, hydro}
#acf detrended plot for hydro
Hydro_acf_detrend= autoplot(Acf(hydro_detrend_ts,lag.max=40))

#pacf detrendedplot for hydro
Hydro_pacf_detrend= autoplot(Pacf(hydro_detrend_ts,lag.max=40))

#plotting comparison with plots from #1 as grid
plot_grid(Hydro_acf, Hydro_acf_detrend,Hydro_pacf, Hydro_pacf_detrend, nrow=4,align= 'h', rel_heights = c(2, 2,2,2))
```


## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Just by looking at the time series and the acf plots, do the series seem to have a seasonal trend? No need to run any code to answer your question. Just type in you answer below.
> The Renewable energy series does not appear to have a seasonal trend (there is no equally spaced pattern that is apparent to me). But the Hydroelectric series does seem to have a seasonal trend. There are equally spaced peaks and troughs. The ACF plot also looks periodic, with cycles repeating every 12 month lag. 

### Q7
Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) the two time series. Ask R to print the summary of the regression. Interpret the regression output. From the results which series have a seasonal trend? Do the results match you answer to Q6?
```{r}
#seasonal dummies for hydroelectric and total renewable
dummies_hydro = seasonaldummy(hydro_detrend_ts)
dummies_renew = seasonaldummy(renew_detrend_ts)

#creating seasonal means model for both
linear_trend_seasonal_hydro <- lm(hydro_detrend_ts ~ dummies_hydro)
  print(summary(linear_trend_seasonal_hydro))
  
linear_trend_seasonal_renew <- lm(renew_detrend_ts ~ dummies_renew)
  print(summary(linear_trend_seasonal_renew))

```
> Hydroelectric Power Consumption has a seasonal trend,with 46.7% of variability in the series described by seasonal component. Renewable energy production does not have a seasonal component, only 3.2% of the variability in the series is described by the seasonal component. This matches my response to question 6. 

### Q8
Use the regression coefficients from Q7 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?
```{r}
# get intercept and slopes and make an equation, subtract from detrended series. 
beta0_season_hydro <- linear_trend_seasonal_hydro$coefficients[1]   # intercept for hydro deseason
beta0_season_renew <- linear_trend_seasonal_renew$coefficients[1]  # intercept for renew deseason
beta1_season_hydro <- linear_trend_seasonal_hydro$coefficients[2:12]  # slope for hydro deseason
beta1_season_renew <- linear_trend_seasonal_renew$coefficients[2:12] # slope for renew deseason

#storing seasonal components
seas_component_hydro=array(0,nobs) 
seas_component_renew= array(0,nobs)

for(i in 1:nobs){
seas_component_hydro[i]=(beta0_season_hydro+beta1_season_hydro%*%dummies_hydro[i,]) 
seas_component_renew[i]=(beta0_season_renew+beta1_season_renew%*%dummies_renew[i,])
}

#subtracting seasonal component from original series
y_deseason_renew <- energy_interest[, 2] - seas_component_hydro #is this correct?!

#Transform into a ts object
ts_seasonal_component_hydro <- ts(seas_component_hydro,start = c(1973,1), frequency = 12)
ts_seasonal_component_renew <- ts(seas_component_renew,start = c(1973,1), frequency = 12)


```


### Q9
Plot ACF and PACF for the deseason series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. nut mot mandatory. Did the plots change? How?
```{r}

```



