---
title: "ENV 797 - Time Series Analysis for Energy and Environment Applications | Spring 2024"
subtitle: "Assignment 6 - Due date 02/28/24"
author: "Emma Kaufman"
output: pdf_document
geometry: margin=2.54cm
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp24.Rmd"). Then change "Student Name" on line 4 with your name.

Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai.

R packages needed for this assignment: "ggplot2", "forecast", "tseries" and "sarima". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(ggplot2)
library(Kendall)
library(lubridate)
library(tidyverse)
library(ggthemes)
#install.packages('sarima')
library(sarima)
library(dplyr)
library(cowplot)
```

This assignment has general questions about ARIMA Models.

## Q1

Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models:

* AR(2)

> Answer: 
ACF: Whether or not the data are autoregressive (significant values at early lags and decay exponentially with time). 
PACF: The order of the model (number of terms) is determined by the number of significant lags in PACF. An AR(2) model will have two signficnat lags in the PACF. 

* MA(1)

> Answer:
ACF: Gives the order of the model. An MA(1) will show one significant lag
PACF: Exponential decay in lags, "short memory" process that is less dependent upon the past. 

## Q2

Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$.

(a) Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use the `arima.sim()` function in R to generate $n=100$ observations from each of these three models. Then, using `autoplot()` plot the generated series in three separate graphs.

```{r ARMA(1,0)}
set.seed(100)
n = 100 
phi <- 0.6 #AR coefficient
theta <- 0.9 #MA coefficient
p <- 1 #order of AR component
q <- 0 #order of MA component

arma1_0 <- arima.sim(model = list(ar = (phi*p), ma = (theta*q)), n)
autoplot(arma1_0)
```

```{r ARMA(0,1)}
p <- 0 #order of AR component
q <- 1 #order of MA component

arma0_1 <- arima.sim(model = list(ar = (phi*p), ma = (theta*q)), n)
autoplot(arma0_1)
```

```{r ARMA(1,1)}
p <- 1 #order of AR component
q <- 1 #order of MA component

arma1_1 <- arima.sim(model = list(ar = (phi*p), ma = (theta*q)), n)
autoplot(arma1_1)
 
```
(b) Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use `cowplot::plot_grid()`).

```{r}
plot_grid(
  autoplot(Acf(arma0_1, lag = 40, plot=FALSE), 
                main = "ACF ARMA(0,1)"),
   autoplot(Acf(arma1_0, lag = 40, plot=FALSE), 
                main = "ACF ARMA(1,0)"),
   autoplot(Acf(arma1_1, lag = 40, plot=FALSE), 
                main = "ACF ARMA(1,1)"),
  ncol=3
  )
```

(c) Plot the sample PACF for each of these models in one window to facilitate comparison.

```{r}
plot_grid(
  autoplot(Pacf(arma0_1, lag = 40, plot=FALSE), 
                main = "PACF ARMA(0,1)"),
   autoplot(Pacf(arma1_0, lag = 40, plot=FALSE), 
                main = "PACF ARMA(1,0)"),
   autoplot(Pacf(arma1_1, lag = 40, plot=FALSE), 
                main = "PACF ARMA(1,1)"),
  ncol=3
  )
```

(d) Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be able identify them correctly? Explain your answer.

> Answer: No we cannot identify them correctly. There is seasonality in both, which makes it difficult to discern whether or not there are AR or MA patterns for a given PACF/ACF.
For the ARMA (0,1) model we would expect to see 1 significant ACF component and an exponentially decaying PACF. 
For the ARMA (1,0) model we would expect to see 1 significant PACF component and an exponentially decaying ACF. 
For the ARMA (1,1) model we would expect.... 

(e) Compare the PACF values R computed with the values you provided for the lag 1 correlation coefficient, i.e., does $\phi=0.6$ match what you see on PACF for ARMA(1,0), and ARMA(1,1)? Should they match?

> Answer:For the ARMA(1,0) we would expect to see a PACF first component with a value 0.6, but the value is 0.35.
For the ARMA(1,1) 

(f) Increase number of observations to $n=1000$ and repeat parts (b)-(e).

```{r}
n= 1000
p <- 1 #order of AR component
q <- 0 #order of MA component
arma1_0_n1000 <- arima.sim(model = list(ar = (phi*p), ma = (theta*q)), n)
autoplot(arma1_0_n1000)

p <- 0 #order of AR component
q <- 1 #order of MA component
arma0_1_n1000 <- arima.sim(model = list(ar = (phi*p), ma = (theta*q)), n)
autoplot(arma0_1_n1000)

p <- 1 #order of AR component
q <- 1 #order of MA component
arma1_1_n1000 <- arima.sim(model = list(ar = (phi*p), ma = (theta*q)), n)
autoplot(arma1_1_n1000)
```

```{r}
#acf for simulations with n=1000
plot_grid(
  autoplot(Acf(arma0_1_n1000, lag = 40, plot=FALSE), 
                main = "ACF ARMA(0,1)"),
   autoplot(Acf(arma1_0_n1000, lag = 40, plot=FALSE), 
                main = "ACF ARMA(1,0)"),
   autoplot(Acf(arma1_1_n1000, lag = 40, plot=FALSE), 
                main = "ACF ARMA(1,1)"),
  ncol=3
  )
```

```{r}
#pacf for simulations with n=1000
plot_grid(
  autoplot(Pacf(arma0_1_n1000, lag = 40, plot=FALSE), 
                main = "PACF ARMA(0,1)"),
   autoplot(Pacf(arma1_0_n1000, lag = 40, plot=FALSE), 
                main = "PACF ARMA(1,0)"),
   autoplot(Pacf(arma1_1_n1000, lag = 40, plot=FALSE), 
                main = "PACF ARMA(1,1)"),
  ncol=3
  )
```


(d.2) Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be able identify them correctly? Explain your answer.

> Answer:

(e.2) Compare the PACF values R computed with the values you provided for the lag 1 correlation coefficient, i.e., does $\phi=0.6$ match what you see on PACF for ARMA(1,0), and ARMA(1,1)? Should they match?

> Answer:

## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$

(a) Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.
> Answer:
p (# of AR terms) = 2?
q (# of MA terms) = 1
a_t (error) = a_t
seasonal terms?! 


(b) Also from the equation what are the values of the parameters, i.e., model coefficients.
> Answer:
phi_1 = 0.7
phi_2 = -0.25
theta_1 = -0.1 


## Q4

Simulate a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with $\phi =0 .8$ and $\theta = 0.5$ using the `sim_sarima()` function from package `sarima`. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot the generated
series using `autoplot()`. Does it look seasonal?

```{r}

```

## Q5

Plot ACF and PACF of the simulated series in Q4. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r}

```
